A novel framework for linking dynamic experiences to memories reveals event-like structure in naturalistic episodic recall

### 2019-07-23 paper/main.tex
finished first pass through revision

@@

% One may reasonably hypothesize that the diminished effect of presentation order on recall likelihood stems from 1) differences between the presence of certain features of episodic experiences that are absent in the simple, discretized stimuli of classic list-learning experiments.  Random word lists most often employ simple, common words While they are well-suited to the random word-list paradigm, these analyses address only the temporal order of stimuli, neglecting to consider their content (or, for example, that content's potential interrelatedness).  This difference serves to highlight the need for alternative approaches to quantifying memory success that account for such contextual features and the way they may impact memory organization and retrieval dynamics.

@@

\paragraph{Temporal clustering scores.} Temporal clustering refers to the extent to which participants group their recall responses according to encoding position~\citep{PolyEtal09}. For instance, if a participant recalled the video events in the exact order they occurred (or in exact reverse order), this would yield a score of 1.  If a participant recalled the events in random order, this would yield an expected score of 0.5.  For each recall event transition (and separately for each participant), we sorted all not-yet-recalled events according to their absolute lag (i.e., distance away in the video).  We then computed the percentile rank of the next event the participant recalled.  We averaged these percentile ranks across all of the participant's recalls to obtain a single temporal clustering score for the participant (mean: 0.808, SEM: 0.022).  Overall, we found that participants with higher temporal clustering scores also tended to recall more events (Pearson's $r(15) = 0.62, p = 0.007$).

\paragraph{Semantic clustering scores.} Semantic clustering measures the extent to which participants clustered their recall responses according to semantic similarity~\citep{PolyEtal09}. Here, we used the topic vectors for each event as a proxy for its semantic content. Thus, the similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For each recall event transition, we sorted all not-yet-recalled events according to how correlated the topic vector \textit{of the closest-matching video event} was to the topic vector of the closest-matching video event to the just-recalled event.  We then computed the percentile rank of the observed next recall.  We averaged these percentile ranks across all of the participant's recalls to obtain a single semantic clustering score for the participant (mean: 0.813, SEM: 0.022).  We found that participants who exhibited stronger semantic clustering scores overall remembered more video events (Pearson's $r(15) = 0.55, p = 0.02$).

% NOTE: Thinking about removing this analysis? More of a sanity check than something novel and cool, and takes up space. Happy to be convinced otherwise

% To quantify the similarity between the video topic trajectory and individual recall topic trajectories, we considered several novel metrics.  First, we tested whether each participant's recall trajectory matched the video trajectory in a general sense. For each participant we filtered the video trajectory to only include the events that the participant remembered.  We then computed the root mean squared difference (RMSD) between the remaining video events and the (closest-matching) recalled events.  For example, if the topic vectors for a participant's recall event topic vectors matched the corresponding video event topic vectors exactly (and in order), the expected RMSD for those events would be 0.  However, if the participant's recall events did not perfectly match the video events, or if they were out of order, then the RMSD would be greater than 0.  To assess the significance of the match between the video and recall trajectories, we carried out a permutation procedure whereby, for each of 10000 repetitions, we circularly shifted the recall trajectories (in time) by a random amount and then re-computed the RMSD each time.  This yielded a distribution of ``null'' RMSD values for each participant.  The observed RMSD values reached significance (i.e., $p < 0.05$, reflecting that more than 95\% of the null RMSD values were greater than the observed RMSD value) for nine of the participants (3, 4, 8--13, and 17).  (For the remaining participants this test yielded $0.05 \leq p < 0.25$.)  The observed RMSD values were also reliably correlated with hand-annotated memory performance across participants (Pearson's $r(15) = -0.57, p = 0.016$).  In other words, a closer match between the video and recall topic trajectories was related to better overall recall performance.

@@

\paragraph*{Precision.}
We tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the correlation between the topic vectors for each recall event and that of its closest-matching video event (only for the events which they recalled). We Fisher's z-transformed the correlations, computed the average and then inverse Fisher's z-transformed the resulting value.  This gave a single value per participant representing the average precision across all recalled events.  We then correlated this value with hand-annotated as well as model derived (e.g. $k$ or the number of events recovered by the HMM) memory performance.

@@

%NOTE: Need to check on this equation!
\[
  d(\mathrm{event}) = 1 - \bar{c}(\mathrm{event}),
\]
where $\bar{c}(\mathrm{event})$ is the average correlation between the given recalled event's topic vector and the topic vectors from all video events \textit{except} the best-matching video event.  We then averaged these distinctiveness scores across all of the events recalled by the given participant. As above, we used Fisher's z (transform and inverse-transform) before/after averaging correlation values. Finally, we correlated these values with hand-annotated and model derived memory performance scores across-subjects.

@@

Next, we constructed two sets of ``template'' matrices: one reflected the video's topic trajectory and the other reflected each participant's recall topic trajectory.  To construct the video template, we computed the correlations between the topic proportions estimated for every pair of TRs (prior to segmenting the trajectory into discrete events; i.e., the correlation matrix shown in Figs.~\ref{fig:model}B and \ref{fig:brainz}A).  We constructed similar temporal correlation matrices for each participant's recall topic trajectory (Figs.~\ref{fig:model}D, \corrmats).  However, to correct for length differences and potential non-linear transformations between viewing time and recall time, we first used dynamic time warping~\citep{BernClif94} to temporally align participants' recall topic trajectories with the video topic trajectory (an example correlation matrix before and after warping is shown in Fig.~\ref{fig:brainz}B).  This yielded a 1976 by 1976 correlation matrix for the video template and for each participant's recall template.

>>>>

We can also apply two analyses native to list-learning studies that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean: 0.808, SEM: 0.022) and that participants with higher temporal clustering scores also tended to recall more events (Pearson's $r(15) = 0.62, p = 0.007$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean: 0.813, SEM: 0.022) and participants who exhibited stronger semantic clustering scores overall remembered more video events (Pearson's $r(15) = 0.55, p = 0.02$).

@@

% NOTE: for the "nearby moments of real life/naturalistic stimuli tend to be semantically similar" argument, is it worth adding something to the effect of:
% 1. we observed high temporal clustering scores
% 2. we also observed high semantic clustering scores
% 3. ergo, for both to be true, temporally proximal video events must be semantically similar
% - Paxton

@@

\paragraph{Temporal clustering scores.} Temporal clustering describes participants' tendency to organize their recall sequences by the learned items' encoding positions.  For instance, if a participant recalled the video events in the exact order they occurred (or in exact reverse order), this would yield a score of 1.  If a participant recalled the events in random order, this would yield an expected score of 0.5.  For each recall event transition (and separately for each participant), we sorted all not-yet-recalled events according to their absolute lag (i.e., distance away in the video).  We then computed the percentile rank of the next event the participant recalled.  We averaged these percentile ranks across all of the participant's recalls to obtain a single temporal clustering score for the participant.

\paragraph{Semantic clustering scores.} Semantic clustering describes participants' tendency to recall semantically similar presented items together in their recall sequences.  Here, we used the topic vectors for each event as a proxy for its semantic content. Thus, the similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For each recall event transition, we sorted all not-yet-recalled events according to how correlated the topic vector \textit{of the closest-matching video event} was to the topic vector of the closest-matching video event to the just-recalled event.  We then computed the percentile rank of the observed next recall.  We averaged these percentile ranks across all of the participant's recalls to obtain a single semantic clustering score for the participant.

% I moved the actual results from the two paragraphs to the results section as an afterthought to the list-learning analyses.  We could also possibly just take them out, though we should keep them if you think it's worth referencing as support for our argument (see my note in discussion section)
% -Paxton

% NOTE: Thinking about removing this analysis? More of a sanity check than something novel and cool, and takes up space. Happy to be convinced otherwise
% leaving this one to Jeremy's judgement
% -Paxton

% To quantify the similarity between the video topic trajectory and individual recall topic trajectories, we considered several novel metrics.  First, we tested whether each participant's recall trajectory matched the video trajectory in a general sense. For each participant, we filtered the video trajectory to only include the events that the participant remembered.  We then computed the root mean squared difference (RMSD) between the remaining video events and the (closest-matching) recalled events.  For example, if the topic vectors for a participant's recall event topic vectors matched the corresponding video event topic vectors exactly (and in order), the expected RMSD for those events would be 0.  However, if the participant's recall events did not perfectly match the video events, or if they were out of order, then the RMSD would be greater than 0.  To assess the significance of the match between the video and recall trajectories, we carried out a permutation procedure whereby, for each of 10000 repetitions, we circularly shifted the recall trajectories (in time) by a random amount and then re-computed the RMSD each time.  This yielded a distribution of ``null'' RMSD values for each participant.  The observed RMSD values reached significance (i.e., $p < 0.05$, reflecting that more than 95\% of the null RMSD values were greater than the observed RMSD value) for nine of the participants (3, 4, 8--13, and 17).  (For the remaining participants this test yielded $0.05 \leq p < 0.25$.)  The observed RMSD values were also reliably correlated with hand-annotated memory performance across participants (Pearson's $r(15) = -0.57, p = 0.016$).  In other words, a closer match between the video and recall topic trajectories was related to better overall recall performance.

@@

\paragraph*{Precision.}
We tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the correlation between the topic vectors for each recall event and that of its closest-matching video event (only for the events which they recalled). We Fisher's $z$-transformed the correlations, computed the average and then inverse Fisher's z-transformed the resulting value.  This gave a single value per participant representing the average precision across all recalled events.  We then correlated this value with hand-annotated as well as model derived (e.g. $k$ or the number of events recovered by the HMM) memory performance.

@@

%NOTE: Need to check on this equation!
\[
  d(\mathrm{event}) = 1 - \bar{c}(\mathrm{event}),
\]
where $\bar{c}(\mathrm{event})$ is the average correlation between the given recalled event's topic vector and the topic vectors from all video events \textit{except} the best-matching video event.  We then averaged these distinctiveness scores across all of the events recalled by the given participant. As above, we used Fisher's $z$-transform (and inverse-transform) before/after averaging correlation values. Finally, we correlated these values with hand-annotated and model derived memory performance scores across-subjects.

@@

Next, we constructed two sets of ``template'' matrices: one reflecting the video's topic trajectory and the other reflecting each participant's recall topic trajectory.  To construct the video template, we computed the correlations between the topic proportions estimated for every pair of TRs (prior to segmenting the trajectory into discrete events; i.e., the correlation matrix shown in Figs.~\ref{fig:model}B and \ref{fig:brainz}A).  We constructed similar temporal correlation matrices for each participant's recall topic trajectory (Figs.~\ref{fig:model}D, \corrmats).  However, to correct for length differences and potential non-linear transformations between viewing time and recall time, we first used dynamic time warping~\citep{BernClif94} to temporally align participants' recall topic trajectories with the video topic trajectory (an example correlation matrix before and after warping is shown in Fig.~\ref{fig:brainz}B).  This yielded a 1976 by 1976 correlation matrix for the video template and for each participant's recall template.
